# Setting up your workspace  {.unnumbered}

## Install R and RStudio

Follow the instructions [here](https://www.youtube.com/watch?v=TFGYlKvQEQ4&t=14s) if you don't have R or RStudio downloaded. Select the CRAN mirror nearest to you (probably Oregon State University). If you have a new apple silicon macbook, make sure to download the version of R that says "Apple silicon arm64 build".

An alternative: R and RStudio are both already installed on [all academic workstations](https://library.uoregon.edu/library-technology-services/public-info/a-software) at UO. The downside is the limited hours, especially on weekends.

## Get Acquainted with the RStudio IDE

 Watch this [video from RStudio](https://www.rstudio.com/products/rstudio/?wvideo=520zbd3tij) to learn a little about the RStudio IDE. Don't get overwhelmed, we'll only use a small subset of the things in there and you'll learn very quickly what's useful to you.

## Install the Tidyverse

Run these lines of code in your console to make sure you have the tidyverse installed and attached to your current session.

```{r}
#| eval: false

install.packages("tidyverse", dependencies = TRUE)
library(tidyverse)
```

## Install gapminder

  You'll use this package a lot in the koans.

```{r}
#| eval: false

install.packages("gapminder")
library(gapminder)
```

## Install a few Packages we'll use for Plots

```{r}
#| eval: false
install.packages("gganimate", dependencies = TRUE)
install.packages("hexbin")
```

## Install `qelp`

 `qelp` (quick help) is an alternative set of beginner friendly help docs I created (with contributions from previous EC421 students) for commonly used functions in R and the tidyverse. Once you have the package installed, you can access the help docs from inside RStudio.

```{r}
#| eval: false
 install.packages("Rcpp", dependencies = TRUE)
 install.packages("devtools", dependencies = TRUE)
 library(devtools)
 install_github("cobriant/qelp")
```

 Now run:

```{r}
#| eval: false
?qelp::install.packages
```

 If everything went right, the help docs I wrote on the function `install.packages` should pop up in the lower right hand pane. Whenever you want to read the qelp docs on a function, you type `?`, `qelp`, two colons `::` which say "I want the help docs on this function which is from the package qelp", and then the name of the function you're wondering about.

## Install the Tidyverse Koans

Visit the [koans on github](https://github.com/cobriant/tidyverse_koans).

Click on the green button that says `Code` and then hit `Download ZIP`.

Find the file (probably in your downloads folder). On Macs, opening the file will unzip it. On Windows, you'll right-click and hit "extract". Then navigate to the new folder named `tidyverse_koans-main` and double click on the R project `tidyversekoans.Rproj`. RStudio should open. If it doesn't, open RStudio and go to `File > Open Project` and then find `tidyversekoans.Rproj`.

In RStudio, go to the lower righthand panel and hit the folder `R`. This takes you to a list of 20 exercises (koans) you'll complete as homework over the course of the quarter. The first 3 (`K01_vector`, `K02_tibble`, and `K03_pipe`) will be due before class on Wednesday (July 20).

Open the first koan: `K01_vector.R`. Before you start, modify 2 keybindings: 

First, make it so that you can hit `Cmd/Ctrl Shift K` to compile a notebook:

`Macs: Tools > Modify keyboard shortcuts > filter for Compile a Notebook > Cmd Shift K`

`Windows: Tools > > Modify keyboard shortcuts > filter for Compile a Notebook > Ctrl Shift K`

Second, make it so that you can hit `Cmd/Ctrl Shift T` to run the test for only the active koan instead of all the koans:

`Macs: Tools > Modify keyboard shortcuts > Run a test file > Cmd Shift T`

`Windows: Tools > Modify keyboard shortcuts > Run a test file > Ctrl Shift T`

Now hit `Cmd/Ctrl Shift T` (Cmd Shift T on a mac; Ctrl Shift T on windows). You've just tested the first koan. You should see:

`[ FAIL 0 | WARN 0 | SKIP 9 | PASS 0 ]`

What does this mean? If there are errors in your R script, the test will not complete. Since it completed, you know there are no errors. Since `FAIL` is 0, you also haven't failed any of the questions yet. But `PASS` is also 0, so you haven't passed the questions either. Since they're blank right now, the test will skip them. That's why `SKIP` is 9.

The tests are meant to help you figure out whether you're on the right track, but they're not perfect: if you keep failing the tests but you think your answer is correct, don't spend too much time worrying about it. The tests are sometimes a little fragile... They're a work in progress!

Go ahead and start working on the koans and learning about the tidyverse! There's no need to wait until they're due to start the koans. I find that the students who end up becoming the strongest programmers spend a lot of time making sure their koans are well done. 

When you're finished with a koan, make sure to run the tests one last time (`Ctrl/Cmd Shift T`) and then publish an html verson of the document (`Ctrl/Cmd Shift K`, and if that doesn't do anything, change the keybinding for `File > Compile Report` to be `Ctrl/Cmd Shift K`). You'll upload the html version to Canvas for me to grade.

One last thing: whenever you want to work on the koans, make sure you open RStudio by opening the `tidyverse_koans-main` project, not just the individ
ual koan file. If you open the koans in a session that's not associated with the `tidyverse_koans-main` project, the tests will fail to run. You can always see which project your current session is being associated with by looking at the upper right hand corner of RStudio: if you're in the `tidyverse_koans-main` project, you'll see `tidyverse_koans-main` up there. That's good. If you're in no project at all, you'll see `Project: (None)` up there. That's not good, especially if you want the tests to run. If you see `Project: (None)`, just click that text and you'll be able to switch over to the `tidyverse_koans-main` project.
